---
title: "Your Project Title (be descriptive and creative!)"
author: "Group 35: Max Ming Yi Koh (1007972785) and Kevin Le (1007952805)"
date: March 31, 2022
output: 
  beamer_presentation:
    theme: "Pittsburgh"
    colortheme: "orchid"
    fonttheme: "structurebold"
    slide_level: 2
classoption: "aspectratio=169"
fontsize: 11pt
urlcolor: blue
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
# echo=FALSE will stop the code chunk from appearing in the knit document
# warning=FALSE and message=FALSE will stop R messages from appearing in the knit document
library(tidyverse)
library(dplyr)
# here is the data for this project
expedia_data <- read_csv("ExpediaSearchData.csv")
glimpse(expedia_data)
# see the Project Overview page for more information: 
# https://q.utoronto.ca/courses/235890/pages/project-overview
```

## Overall Introductions

## Data Summary

## Research Question 1
Does urgency affect how much consumers are willing to pay? 

--- 
# TODO: Add visualization 
```{r}
expedia_data_q1 <- expedia_data %>% 
  select(is_trans1, is_trans2, is_trans3) %>% 
  mutate(is_trans1 = ifelse(!is.na(is_trans1) & is_trans1 == 1, 1, 0),
         is_trans2 = ifelse(!is.na(is_trans2) & is_trans2 == 1, 1, 0), 
         is_trans3 = ifelse(!is.na(is_trans3) & is_trans3 == 1, 1, 0),
         trans_made = ifelse(is_trans1 + is_trans2 + is_trans3 > 0, TRUE, FALSE))


# plot a bar chart 9 vs 991


# expedia_data_q1 <- expedia_data %>% 
#   select(is_trans1, is_trans2, is_trans3, price_bucket1, price_bucket2, price_bucket3, 
#          checkin_date, search_timestamp, destination_id) %>% 
#   filter(is_trans1 + is_trans2 + is_trans3 == 1 & 
#            !is.na(checkin_date) & !is.na(search_timestamp) & !is.na(destination_id)) %>%
#   mutate(
#     price_bucket1 = ifelse(is.na(is_trans1) | is_trans1 == 0, 0, price_bucket1),
#     price_bucket2 = ifelse(is.na(is_trans2) | is_trans2 == 0, 0, price_bucket2),
#     price_bucket3 = ifelse(is.na(is_trans3) | is_trans3 == 0, 0, price_bucket2),
#     price_bucket = (price_bucket1 + price_bucket2 + price_bucket3) / (is_trans1 + is_trans2 + is_trans3),
#     search_timestamp = as.Date(as.character(search_timestamp)))
# expedia_data_q1 %>% ggplot(aes(x=checkin_date - search_timestamp, y=price_bucket)) + 
# 	geom_point() + 
# 	labs(x = "TODO: add x-axis label", 
# 		   y = "TODO: add y-axis label", 
# 		   title = "TODO: add title") + 
# 	geom_smooth(method="lm", se=FALSE) + 
# 	theme_minimal() 
```

--- 
# TODO: Set up statistical model 
```{r}
set.seed(15)
sample_means <- rep(NA, 1000) 

for(i in 1:1000) { 
	# sample for new 25 rows each iteration
  sample <- expedia_data_q1 %>% sample_n(size = 1000, replace = TRUE)
  sample_means[i] <- as.numeric(sample %>% summarize(mean(trans_made)))
}

# convert into a tibble
sample_means = tibble(trans_proportion = sample_means) 

quantile(sample_means$trans_proportion, c(0.025, 0.975))
# expedia_data_q2 <- expedia_data %>% 
#   select(is_trans1, is_trans2, is_trans3, price_bucket1, price_bucket2, price_bucket3, 
#          checkin_date, search_timestamp, destination_id) %>% 
#   filter(is_trans1 + is_trans2 + is_trans3 == 1 & 
#            !is.na(checkin_date) & !is.na(search_timestamp) & !is.na(destination_id)) %>%
#   mutate(
#     price_bucket1 = ifelse(is.na(is_trans1) | is_trans1 == 0, 0, price_bucket1),
#     price_bucket2 = ifelse(is.na(is_trans2) | is_trans2 == 0, 0, price_bucket2),
#     price_bucket3 = ifelse(is.na(is_trans3) | is_trans3 == 0, 0, price_bucket2),
#     price_bucket = (price_bucket1 + price_bucket2 + price_bucket3) / (is_trans1 + is_trans2 + is_trans3),
#     search_timestamp = as.Date(as.character(search_timestamp)))
# expedia_data_q1 %>% ggplot(aes(x=checkin_date - search_timestamp, y=price_bucket)) + 
# 	geom_point() + 
# 	labs(x = "TODO: add x-axis label", 
# 		   y = "TODO: add y-axis label", 
# 		   title = "TODO: add title") + 
# 	geom_smooth(method="lm", se=FALSE) + 
# 	theme_minimal() 
```

--- 
# TODO: Analyze results from statistical model 
- we are 95% confident that the number of people who transact one of the top 3 listings on Expedia is (0.003, 0.015). This means that if we were to repeat the experiment ...  

## Research Question 2 
Are properties with discounted prices more likely to be sold? 

Motivation of research question: 

--- 
# TODO: Add visualization 

--- 
# TODO: Set up statistical model 
```{r}
untidy_expedia_data_q2 <- expedia_data %>% 
  select(num_clicks1, is_drr1, num_clicks2, is_drr2, num_clicks3, is_drr3)

expedia_data_q2 <- tibble(num_clicks = c(NA), is_drr = c(NA))
add_row(expedia_data_q2, num_clicks = 1, is_drr = 2) 
add_row(expedia_data_q2, num_clicks = 3, is_drr = 4)
expedia_data_q2

# 
# for (i in 1:nrow(untidy_expedia_data_q2)) {
#   add_row(expedia_data_q2, num_clicks = untidy_expedia_data_q2[i,"num_clicks1"], is_drr = untidy_expedia_data_q2[i,"is_drr1"])
#   add_row(expedia_data_q2, num_clicks = untidy_expedia_data_q2[i,"num_clicks2"], is_drr = untidy_expedia_data_q2[i,"is_drr2"])
#   add_row(expedia_data_q2, num_clicks = untidy_expedia_data_q2[i,"num_clicks2"], is_drr = untidy_expedia_data_q2[i,"is_drr2"])
# }
# 
# expedia_data_q2 %>% 
#   filter(!is.na(num_clicks) & !is.na(is_drr))
# 
# expedia_data_q2
```

--- 
# TODO: Analyze results from statistical model 

## Research Question 3 
Do the number of children and adults affect the length of travel?

--- 
# TODO: Add visualization 
```{r}
expedia_data_q3 <- expedia_data %>% 
  select(search_timestamp, adult_count, child_count, checkin_date) %>%
  filter(!is.na(checkin_date) & !is.na(search_timestamp)) %>% 
  mutate(travel_length = as.numeric(as.Date(as.character(checkin_date)) - as.Date(as.character(search_timestamp))))

glimpse(expedia_data_q3)

expedia_data_q3 %>% ggplot(aes(x=child_count, y= travel_length)) +
  geom_point() +
  labs(x="number of children",
       y="length of travel (days)",
       title = "TODO: add title") +
  theme_minimal()

expedia_data_q3 %>% ggplot(aes(x=adult_count, y= travel_length)) +
  geom_point() +
  geom_smooth(se=FALSE, method="lm") +
  labs(x="number of adults",
       y="length of travel (days)",
       title = "TODO: add title") +
  theme_minimal()

# TODO: add model 3 and 4 
# model 3's starter code 
mariokart2 %>% ggplot(aes(x = duration, y = total_pr, colour = cond)) + 
	labs(x = "Duration (in days)", y = "Total Price") + 
	geom_point() 

# model 4's starter code 
library(broom)
augment(model3) # creates dataframe with predicted value in .fitted column 
# (y_i hat). There is 1 row for each observation in the training data 

mariokart2 %>% ggplot(aes(x = duration, y = total_pr, color = cond)) + 
	geom_point(alpha = 0.5) +     # alpha controls transparency of points 
	geom_line(data = augment(model3), aes(y = .fitted, colour = cond)) + 
	labs(x = "Duration (in days)"), y = "Total price ($)")
```

--- 
# TODO: Set up statistical model 
```{r}
# 4 models (adult vs length, children vs length, adult + children vs length, adult * children vs length)

set.seed(12) 
n <- nrow(expedia_data_q3)
training_indices <- sample(1:n, size=round(0.8 * n)) 

train <- expedia_data_q3 %>% filter(rowid %in% training_indices) 
y_train <- train$travel_length 

test <- expedia_data_q3 %>% filter(!(rowid %in% training_indices))
y_test <- test$travel_length 

mod1_train <- lm(travel_length ~ adult_count, data = train)
mod2_train <- lm(travel_length ~ child_count, data = train)
mod3_train <- lm(travel_length ~ adult_count + child_count, data = train)
mod4_train <- lm(travel_length ~ adult_count * child_count, data = train)

yhat_mod1_test <- predict(mod1_train, newdata = test)
yhat_mod2_test <- predict(mod2_train, newdata = test) 
yhat_mod3_test <- predict(mod3_train, newdata = test) 
yhat_mod4_test <- predict(mod4_train, newdata = test) 

yhat_mod1_train <- predict(mod1_train, newdata = train)
yhat_mod2_train <- predict(mod2_train, newdata = train) 
yhat_mod3_train <- predict(mod3_train, newdata = train) 
yhat_mod4_train <- predict(mod4_train, newdata = train) 

mod1_test_RMSE <- sqrt(sum((y_test - yhat_mod1_test)^2 / nrow(test)))
mod2_test_RMSE <- sqrt(sum((y_test - yhat_mod2_test)^2 / nrow(test)))
mod3_test_RMSE <- sqrt(sum((y_test - yhat_mod3_test)^2 / nrow(test)))
mod4_test_RMSE <- sqrt(sum((y_test - yhat_mod4_test)^2 / nrow(test)))

mod1_train_RMSE <- sqrt(sum((y_train - yhat_mod1_train)^2 / nrow(train)))
mod2_train_RMSE <- sqrt(sum((y_train - yhat_mod2_train)^2 / nrow(train)))
mod3_train_RMSE <- sqrt(sum((y_train - yhat_mod3_train)^2 / nrow(train)))
mod4_train_RMSE <- sqrt(sum((y_train - yhat_mod4_train)^2 / nrow(train)))

my_table <- tibble(Model = c("1", "2", "3", "4"), 
	                 RMSE_testdata = c(mod1_test_RMSE, mod2_test_RMSE, 
		                                 mod3_test_RMSE, mod4_test_RMSE), 
                   RMSE_traindata = c(mod1_train_RMSE, mod2_train_RMSE, 
                                      mod3_train_RMSE, mod4_train_RMSE),
                   ratio_of_RMSEs = RMSE_testdata / RMSE_traindata)
my_table
summary(mod1_train)$coefficients
summary(mod2_train)$coefficients 
summary(mod3_train)$coefficients 
summary(mod4_train)$coefficients 
```

--- 
# TODO: Analyze results from statistical model 
- mention that 4 models are tested
- talk about model 1 being the best 
  * simple (in terms of # of predictors) 
  * RMSE ratio fairly similar in all 4 models (no overfitting) 
  * RMSE ratio for test data is fairly similar (equally accurate) 
- based on p-value of 6.675124e-07, there is very strong evidence against the null hypothesis that ... 
- based on the slope of 9.52, it means that on average for each increase in adult in a trip, the trip's length increases by 9.52 days. 


- variation of data is weak as 90% of the dataset are with searchs with no children/ infants
- p-value for the child_count model is approximately 0.563 which suggests that there is weak evidence against the null hypothesis 
- p-value for the adult_count model is relatively small (p-value < 5.393e-08) which suggests very strong evidience against the null hypothesis
- p-value for the infant_count model is relatively small (p-value < 1.393e-02) which suggests strong evidience against the null hypothesis
   ~ more onto the infant sample, it is noticed that only 2 searches have infants in their group which causes small generalizations to be yield here
- all R^2 values (coefficient of determination)  yields a very small value for all three of the models which could be due to the fact of low variation in the data
- we can conclude that the number of adults will lead to greater travel lengths


## Limitations 
- assume that the data filled in is accurate 
- sample is representative of population (1000 people)

## Overall Conclusion
